---
title: "R review week 5"
output: html_notebook
---

# R review

Review of R data types and functions.

## R Data types

### Vectors

Basic type of data in R. Some of the main features of vectors:

- store numbers and characters in one dimension.
- can only store same type of data. If you try to mix it may not warn you and simply pick a type like character.
- use **c()** to create a vector


```{r}
a = c(1,2,3)
b = c("a","b","c")
c = c(a,b)
c
```
#### Logical vectors

```{r}
filter = c(T,T,F)
a < 2
```
To get the position where the vector is T you can use the **which()** command.

```{r}
which(filter)
```
```{r}
which(a<2)
```
#### How to get and set values in vector?

To get a value from a vector you can specify :

- indices,
- range of indices
- names
- logical vector which returns all indices are T

```{r}
a[3]
```
```{r}
a[2:3]
```
```{r}
names(a) = c("apple","banana","orange")
```

```{r}
a["apple"]
```

```{r}
a[filter]
```


### Matrix

Some properties:

- 2 dimensional array
- same type, can be either all numeric, character, or even logical
- can be created using **matrix()** command.

```{r}
mat = matrix(1:20, nrow=5, ncol=4)
mat
```
```{r}
mat[2,]
```
```{r}
mat[2,2]
```
#### Matrix have names as well

```{r}
colnames(mat) = c("a","b","c","d")
rownames(mat) = c("g1","g2","g3","g4","g5")
mat
```
```{r}
mat["g2",]
```
```{r}
mat[c(T,T,F,T,F),]
```
#### Specials functions for Matrix

```{r}
colSums(mat)
```

```{r}
rowSums(mat)
```
```{r}
mat[rowSums(mat) > 40,]
```

##### cbind() and rbind() to create new matrix
```{r}
a2 = 7:9
a_mat = cbind(a,a2)
a_mat
```
```{r}
a_mat_2 = rbind(a,a2)
a_mat_2
```


### Data frames

Some features of a dataframe

- 2 dimension
- more than one type, but they must be column-wise.
- shares some functions as matrix
- columns can also be referred using *$*
- when you select a column, the result is a vector
- when you select a row, the result is a data frame.

```{r}
adf = data.frame(a=a,b=b,a2=a2)
adf
```

```{r}
adf["apple","b"]
```
```{r}
class(adf$b)
```
```{r}
class(adf["apple",])
```
##### converting data frame to matrix

make sure the columns you select are of same type

```{r}
as.matrix(adf[,c(1,3)])
```
##### converting  matrix to data frame 

```{r}
as.data.frame(mat)
```

#### Lists

Lists are useful in creating an object where different types of data can be linked.

Some features of lists:

- one dimensional
- each element can be of any size
- access named elements using *$*
- access unnamed elements using *[[]]*

```{r}
alist = list(a =a, b= b, c = c, a2)
```


```{r}
alist$a
```

```{r}
alist[[4]]
```
```{r}
homework = list(netid='msk8',
                grade=70,
                result = matrix(10,2,5))
```

```{r}
homework$netid
```
```{r}
homework$result
```

