---
title: "limma"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=F}
# loading libraries
library(edgeR)
library(limma)
library(dplyr)
```

```{r}
# storing the count data obtained by salmon in a variable
data = tx.salmon$counts
#dim(data)
```


```{r}
# creating DGE object using egdeT
DGE = DGEList(counts=data)
dim(DGE)
```

```{r}
# loading sample information
info = read.table("info.tsv", sep = '\t', header=T)
head(info)
```

```{r}
shock = substr(info$Col_names, 1, nchar(info$Col_names)-6)
shock
```

```{r}
time = substr(info$Col_names, nchar(info$Col_names)-1, nchar(info$Col_names))
time
```

```{r}
group = interaction(shock,time)
group
```

```{r}
grp = as.data.frame(rbind(shock,time))
grp = as.data.frame(t(grp))
head(grp)
rownames(grp) = rownames(group)
```

```{r}
# creating the design matrix
design.matrix = model.matrix(~0+group)
design.matrix
```
```{r}
# filtering lowly expressed genes
keep = filterByExpr(DGE, design.matrix)
DGE = DGE[keep,,keep.lib.sizes=FALSE]
```


```{r}
dim(DGE$counts)
```

## Voom: Limma
```{r}
#jpeg("voom.jpg")
v = voom(DGE, design.matrix, plot=TRUE, normalize="quantile")
#dev.off()
```

```{r}
# fitting our design matrix to linear model
fit = lmFit(v, design.matrix)

# emperical bayes of fitted contrast
fit = eBayes(fit)
```

```{r}
dim(fit)
```


```{r}
#head(fit)
```


```{r}
# top table of differentially expressed genes 
top.table = topTable(fit, adjust.method = "BH", sort.by = "none", n = Inf)
head(top.table, 20)
```

```{r}
dim(top.table)
```


```{r}
#topTable(fit, coef=ncol(design.matrix))
```


```{r}
# plotting MDS to see the how the groups are clustering

#jpeg("mds.jpg")
plotMDS(v, labels=paste(shock,time, sep="|"))
#dev.off()
```

### 1) testing within shock groups at different time points
```{r}
contrasts_times = makeContrasts(CST2_T1 = groupCS.T2 - groupCS.T1,
                          CST3_T1 = groupCS.T3 - groupCS.T1,
                          CST3_T2 = groupCS.T3 - groupCS.T2,
                          SST2_T1 = groupSS.T2 - groupSS.T1,
                          SST3_T1 = groupSS.T3 - groupSS.T1,
                          SST3_T2 = groupSS.T3 - groupSS.T2,
                          levels = colnames(fit))

contrasts_times
```

```{r}
# fitting the contrast for within time groups
fit4= contrasts.fit(fit, contrasts_times)
```

```{r}
fit4 = eBayes(fit4)

top.table4 = topTable(fit4,  adjust="BH", n=Inf)
head(top.table4, 20)
```

We did not observe any signifcant genes with adjusted p.value < 0.05 

### 2) between shock groups at same time points
```{r}
contrasts = makeContrasts(T_1 = groupCS.T1 - groupSS.T1,
                          T_2 = groupCS.T2 - groupSS.T2,
                          T_3 = groupCS.T3 - groupSS.T3,
                          levels = colnames(fit))

contrasts
```


```{r}
fit3= contrasts.fit(fit, contrasts)
```

```{r}
# emperical bayes of fitted contrast
fit3 = eBayes(fit3)
top.table3 = topTable(fit3,  adjust="BH", n=Inf)
head(top.table3, 100)
```
```{r}
length(which(top.table3$adj.P.Val < 0.05))
```


```{r}
length(which(top.table3$adj.P.Val < 0.01))
```

```{r}
volcanoplot(fit3, "T_3", highlight = 10)
```

We observed 461 genes with adjusted p.value < 0.05 and 140 genes with adjusted 
p. value < 0.01


```{r}
# checking for upregulated and down regulated genes
summary(decideTests(fit3))
```



```{r}
# subset of differentially expressed genes which are significant 
# at adj.P.val < 0.05
diffexpgenes = subset(top.table3, adj.P.Val < 0.05)
```

```{r}
# normalization of counts using Counts Per Million function
nc <- cpm(DGE, normalized.lib.sizes=FALSE)
```

```{r}
# subset of diff. genes based on  normalized counts
diffexpgenes_df = as.data.frame(diffexpgenes)
diffexpvalues = nc[rownames(nc) %in% rownames(diffexpgenes_df), ]
```


```{r}
colnames(diffexpvalues) = info$Sample.ID
#head(diffexpvalues)
```

```{r}
group = as.data.frame(group)
```

### Hierarchical clustering
```{r}
dist = dist(diffexpgenes_df)
tree = hclust(dist)
clusters = cutree(tree,k = 8)
clusters_df = as.data.frame(clusters)

group_by(clusters_df, clusters) %>% summarize(count=n()) 
```


```{r}
readcount = tx.salmon$counts
colnames(readcount) = info$Sample.ID
#head(readcount)
```

```{r}
group = as.data.frame(group)
```

```{r}
expgroup = as.data.frame(colnames(diffexpvalues))
dim(expgroup)
```
```{r}
rownames(grp) = rownames(expgroup)
dim(grp)
```


```{r}
expgroup$shock_group = c(shock)
rownames(expgroup) = expgroup$`colnames(diffexpvalues)`
expgroup$`colnames(diffexpvalues)`<- NULL
#head(expgroup)
```


```{r eval=F}
#library('pheatmap')
#pheatmap(diffexpvalues, scale = 'row', cluster_rows = TRUE, annotation_col= expgroup, annotation_row= clusters_df)
```

## pearson's correlation
```{r}
# Pairwise correlation between samples (columns)
cols.cor <- cor(diffexpvalues, use = "pairwise.complete.obs", method = "pearson")
# Pairwise correlation between rows (genes)
rows.cor <- cor(t(diffexpvalues), use = "pairwise.complete.obs", method = "pearson")

# Plot the heatmap
library("pheatmap")

#jpeg("pheatmap.jpg")

pheatmap(
  diffexpvalues, scale = "row", 
  clustering_distance_cols = as.dist(1 - cols.cor),
  clustering_distance_rows = as.dist(1 - rows.cor),
  cluster_rows = TRUE, 
  annotation_col= grp, 
  annotation_row= clusters_df,
  show_rownames = F
  )

#dev.off()

```

```{r}
head(clusters_df)
```

### GO term analysis for differentially expressed genes between shock groups

```{r eval=F}
library('GOstats')
library('Category')
library("org.Hs.eg.db")
library('annotate')
library('GO.db')
```



```{r}
# creating a list of parameters to be used in hyperGtest 
params = new('GOHyperGParams',
             geneIds = diff_genes$`NCBI gene (formerly Entrezgene) ID`, 
             universeGeneIds = data_genes$`NCBI gene (formerly Entrezgene) ID`, 
             annotation = "org.Hs.eg.db", 
             ontology = 'BP', 
             pvalueCutoff = 0.001, 
             testDirection = 'over')

# creating hypergeometric object
overBP = (hyperGTest(params))
```

```{r}
summary(overBP)
```
```{r}
goDag(overBP)
```

```{r}
length(overBP@geneIds)
```

```{r}
# fetching gene IDs for GO Ids
GO_genes = geneIdsByCategory(overBP)
GO0002376 = GO_genes$`GO:0002376`
GO0002376 = na.omit(GO0002376)
GO0002376 = as.character(GO0002376)
```

```{r}
GO0002376_ids = getSYMBOL(GO0006334, data='org.Hs.eg')
```

```{r}
GO0006334.s = GO_genes$`GO:0006334`
GO0006334.s = na.omit(GO0006334.s)
GO0006334.s = as.character(GO0006334.s)
GO0006334.s_ids = getSYMBOL(GO0006334.s, data='org.Hs.eg')
```

```{r}
GO0006959 = GO_genes$`GO:0006959`
GO0006959 = na.omit(GO0006959)
GO0006959 = as.character(GO0006959)
GO0006959_ids = getSYMBOL(GO0006959, data='org.Hs.eg')
GO0006959_ids
```

```{r}
# GO analysis using goana() of limma
g = goana(rownames(fit3), coef=3, species="Hs", geneid = "EnsembleID")
topGO(g, ontology = "BP")
```


```{r}
colnames(fit3)
```


```{r}
head(top.table3)
```

```{r}
topTable(fit3, coef=3, n=Inf)
```

```{r}
# venn diagram representing the results of up and down genes 
res_s = decideTests(fit3)
vennDiagram(res_s, include = c("up", "down"))
summary(res_s)

```

